---
title: "project"
format: html
editor: visual
---

Loading the required libraries

```{r}
library(tidyverse)
library(fpp2)
library(urca)
```

# Part 1 - Investigating global temperatures over time

First, we load in temperature change data. Our goal here is to investigate if temperature change pre 1910s and post 1910s has seen a significant difference. If yes, it suggests that anthropogenic activities have led to climate change.

```{r}
temp_change_df<- read_csv("Data/annual_temp_change.csv")
head(temp_change_df)
tail(temp_change_df)
```

At first glance, we see that there are 2 methods of temperature recording (GCAG and CISTEMP). We shall evaluate both temp series time series data separately.

```{r}
temp_change_GCAG <- temp_change_df |>
  filter(Source=="GCAG")|>
  arrange(Year)
```

```{r}
# Convert to ts object
temp_change_GCAG <- ts(temp_change_GCAG["Mean"],
                 start=c(1882), 
                 end=c(2016))

autoplot(temp_change_GCAG)
```

```{r}
temp_change_GISTEMP <- temp_change_df |>
  filter(Source=="GISTEMP")|>
  arrange(Year)

# Convert to ts object
temp_change_GISTEMP <- ts(temp_change_GISTEMP["Mean"],
                 start=c(1882), 
                 end=c(2016))

autoplot(temp_change_GISTEMP)
```

Both have consistent results. Do we stick to one or both? There is clear trend. Seasonality might need to decompose to see or use the ACF. Thank you.

```{r}


```
