---
title: "Time Series HW 2"
format: html
editor: visual
---

Load required packages:

```{r}
library(fpp2)
library(forecast)
library(tidyverse)
library(TSstudio)
```
## 1a
```{r}


```

## 3a

```{r}
autoplot(advert, facets=TRUE)

```

It use useful to set facets=TRUE, since there are 2 variables with different scale and magnitude. Hence, a facet plot would be able to effectively illustrate this better than a combined plot.

## 3b

```{r}
tslm(sales~advert, advert)

```

## 3c

To show that the residuals have significant autocorrelation,

```{r}
checkresiduals(tslm(sales~advert, advert))
```

The p value of 0.02856 \< 0.05, hence we rej H0 (there is no autocorrelation) to conclude there is autocorrelation

## 3d

```{r}
Arima(advert[,"sales"], xreg=advert[,"advert"],
  order=c(0,0,0))
checkresiduals(Arima(advert[,"sales"], xreg=advert[,"advert"],
  order=c(0,0,0)))
```

No difference, still fails ljung box test (p value 0.0002\<0.05)

## 3e & f

```{r}
auto.arima(advert[,2], xreg=advert[,"advert"], lambda="auto")
checkresiduals(advert[,2], xreg=advert[,"advert"], lambda="auto")
```

In this case, auto.arima() estimates ARIMA(0,1,0).The model passes the ljung box test as p value 0.59\>0.05.

## 3g

```{r}
forecast(auto.arima(advert[,2], xreg=advert[,"advert"], lambda="auto"), h=6, xreg = rep(10,6))
autoplot(forecast(auto.arima(advert[,2], xreg=advert[,"advert"], lambda="auto"), h=6, xreg = rep(10,6)))
```

## 4a & b

```{r}
set.seed(123)
y <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y[i] <- 0.6*y[i-1] + e[i]
```

As I change phi 1, the coefficient on Y~t-1~ changes, hence the time plot changes. phi 1 has to be between -1 and 1, for the time plot to be legit. Outside this range, the time plot is no longer stationary.

## 4c & d

Creating an MA(1) with theta as 0.6

```{r}
y <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y[i] <- 0.6*e[i-1] + e[i]
```

Changing around the theta value

```{r}
y <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y[i] <- -1.6*e[i-1] + e[i]

autoplot(y)
```

As I change theta 1, the coefficient on e~t-1~ changes, hence the time plot changes.

## 4e

Generating an ARMA(1,1) with coefficients 0.6

```{r}
y <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y[i] <- 0.6*y[i-1] + 0.6*e[i-1] + e[i]
```

## 4f

Generating an AR(2) with phi1 = -0.8 and phi2 = 0.3

```{r}
z <- ts(numeric(100))
e <- rnorm(100)
for(i in 3:100) {
  z[i] <--0.8*z[i-1] + 0.3*z[i-2] + e[i]
}
```

## 4g

Graphing the 2 series and comparing them

```{r}
autoplot(y)
autoplot(z)
```

Comparing the first plot to the second, the first plot exhibits stationarity while the second plot has a constant mean but increasing variance.

## 5
```{r}
df <- read_csv("C:/Users/rayne/Documents/GitHub/DSA301_TimeSeries/data/output_timeseries.csv")
singers_ts <- ts(df, start=c(2020,1), end=c(2023,12), deltat=1/365)

```
Splitting training/test set to 80 20 ratio

```{r}
outofsample= 292
singers_ts_split <- ts_split(singers_ts, sample.out=outofsample)

```

ARIMA to forecast taylor swift
```{r}
arima1 <- auto.arima(singers_ts_split$train[,2], lambda="auto")
arima_fc <- forecast(arima1, h=outofsample)
accuracy(x=singers_ts_split$test[,2], arima_fc)

```
ARIMA x to forecast taylor swift with Justin bieber as eXogenous
```{r}
# taylor swift
# Parameterising xreg so that xreg can be editted to include as many exogenous actors as needed
xreg=singers_ts_split$train[,3] # Justin Bieber

# Build ARIMA x and test model accuracy
arimax1 <- auto.arima(singers_ts_split$train[,2], lambda="auto", xreg=xreg)
arimax1_fc <- forecast(arimax1, h=outofsample, xreg=xreg)
accuracy(x=singers_ts_split$test[,2], arimax1_fc)
```

ARIMA x to forecast taylor swift with Justin bieber, Selena Gomez, Rihanna as eXogenous
```{r}
# taylor swift
# Parameterising xreg so that xreg can be editted to include as many exogenous actors as needed
xreg=singers_ts_split$train[,3:5] # Justin Bieber, Selena Gomez, Rihanna

# Build ARIMA x and test model accuracy
arimax2 <- auto.arima(singers_ts_split$train[,2], lambda="auto", xreg=xreg)
arimax2_fc <- forecast(arimax2, h=outofsample, xreg=xreg)
accuracy(x=singers_ts_split$test[,2], arimax2_fc)
```