---
title: "ts_project"
format: html
editor: visual
---

Loading required libraries

```{r}
library(tidyverse)
library(fpp2)
library(urca)
```

# 1.0 Green model (Temperature, Sea levels and CO2 emissions)

## 1.1 Data processing

Loading and cleaning global temperature dataset from Berkley Earth.

```{r}
global_temp_df <- read_csv("C:/Users/rayne/Documents/GitHub/DSA301_TimeSeries/Data/GlobalTemperatures.csv")

# Remove NA values in LandAndOceanAverageTemperature
global_temp_df <- global_temp_df |>
  filter(!is.na(LandAndOceanAverageTemperature))

# Converting from monthly to yearly, then renaming columns for consistency
global_temp_df <- global_temp_df |>
  group_by(year(dt)) |>
  summarize(LandAverageTemperature=mean(LandAverageTemperature))|>
  rename(year=`year(dt)`, global_temperature=LandAverageTemperature)

# range(global_temp_df$year)
```

Loading and cleaning sea levels data

```{r}
sea_levels_df <- read_csv("C:/Users/rayne/Documents/GitHub/DSA301_TimeSeries/Data/Global Mean Sea Levels (1880 to 2013).csv")

# Selecting only relevant columns, then renaming for consistency
sea_levels_df <- sea_levels_df |>
  select(Year, `CSIRO Adjusted Sea Level`)|>
  rename(year=Year, sea_level=`CSIRO Adjusted Sea Level`)
```

Loading and cleaning carbon emissions data
```{r}
emissions_df <- read_csv("C:/Users/rayne/Documents/GitHub/DSA301_TimeSeries/Data/emission data.csv")

# Pivot longer to make years become rows instead of columns in the raw data, then filtering World to get the global emissions, then selecting only the relevant columns, and lastly converting year from character to numeric for left join later on.
emissions_df <- emissions_df |>
  pivot_longer(-Country, names_to="year", values_to="carbon_emissions")|>
  filter(Country=="World", year>1879)|>
  select(-Country)|>
  mutate(year=as.numeric(year))
```
Loading world population data, with data pre-1951s naively filled. In the past, population did not grow as quickly, which justifies the use of a naive method for filling. 

Population data is to help normalize CO2 emissions as part of feature engineering, in order to improve green model performance.
```{r}
world_pop <- read_csv("C:/Users/rayne/Documents/GitHub/DSA301_TimeSeries/Data/world_population.csv")

world_pop <- world_pop |>
  filter(Year<2018)|>
  select(year=Year, `World Population`)|>
  arrange((year))
```
Now, we get the emissions per global capita.
```{r}
emissions_df <-emissions_df |>
  mutate(carbon_emissions_per_capita=emissions_df$carbon_emissions/world_pop$`World Population`)|>
  select(year, carbon_emissions_per_capita, carbon_emissions)
```

Merging the 3 cleaned datasets to a green_model_df, which will be used to create models for the the green model.
```{r}
green_model_df <-global_temp_df |>
  left_join(sea_levels_df, by="year")|>
  left_join(emissions_df, by="year")

```

For the complete green_model_df, we start from 1881 to 2013 (sea_level_df is the limiting dataset). Lastly, we have to convert year from numeric to date.
```{r}
green_model_df <-green_model_df |>
 filter(year>1879 & year<2014)|>
  mutate(year=make_date(year))

```

Creating timeseries object with the 3 variables for multivariate modelling, and autoplotting each respective variable for data visualization.
```{r}
# Create TS
green_model_ts <- ts(green_model_df[c("carbon_emissions_per_capita", 
                                      "global_temperature", 
                                      "sea_level")],
                     start=c(1880,1),
                     end=c(2013,1))

green_model_ts |> autoplot()

ts(green_model_df[c("global_temperature")],
                     start=c(1880,1),
                     end=c(2013,1)) |> autoplot()

ts(green_model_df[c("carbon_emissions")],
                     start=c(1880,1),
                     end=c(2013,1))|> autoplot()

ts(green_model_df[c("sea_level")],
                     start=c(1880,1),
                     end=c(2013,1))|> autoplot()
```

## 1.2 Model building

Building VAR models
```{r}
vars::VARselect(green_model_ts[,1:3])

```
Check if VAR model passes tests
```{r}
var1 <- vars::VAR(green_model_ts[,1:3], p=1, type="const")
vars::serial.test(var1, lags.pt=10, type="PT.asymptotic")
```

```{r}
var2 <- vars::VAR(green_model_ts[,1:3], p=2, type="const")
vars::serial.test(var2, lags.pt=10, type="PT.asymptotic")
```

```{r}
var3 <- vars::VAR(green_model_ts[,1:3], p=3, type="const")
vars::serial.test(var3, lags.pt=10, type="PT.asymptotic")

```
Now we measure the goodness-of-fit by looking at out-of-sample RMSE and MAPE for the various VAR models.

VAR of order 1
```{r}
outofsample=10
green_model_ts_split <- ts_split(green_model_ts, sample.out=outofsample)
var3 <- vars::VAR(green_model_ts_split$train[,1:3], p=1, type="const")
var_fc <- forecast(var3, h=outofsample)
accuracy(x=green_model_ts_split$test[,2:2], var_fc$forecast$global_temperature)
```
VAR of order 3
```{r}
outofsample=10
green_model_ts_split <- ts_split(green_model_ts, sample.out=outofsample)
var3 <- vars::VAR(green_model_ts_split$train[,1:3], p=3, type="const")
var_fc <- forecast(var3, h=outofsample)
accuracy(x=green_model_ts_split$test[,2:2], var_fc$forecast$global_temperature)
```

